---
title: "Project_1_Main"
author: "Michael Arango, Yeshwant Chillakuru, Jackson Crum"
date: "February 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Data cleaning and munging packages
library(dplyr)
library(tidyverse)
library(tidyr)
library(doBy)
# date format library
library(lubridate)
# geolibraries
library(rgdal)
library(sp)
library(raster)
# mapping/plotting
library(ggmap)
library(ggplot2)
library(foreign)
```

#Description of the Dataset

##Reading the Data into R
```{r Reading in Data, tidy=TRUE, results='hide'}
getwd()
tickets <- data.frame(read.csv('Muni_Driver_Reported_Transit_Only_Lane_Violations.csv', header = T, sep = ","))
```

##Checking Dimensions and Structure of the Dataset 
```{r Variables, tidy=T, results='hide'}
#Check dimensions of dataset 
dim(tickets)
#Check variable names
colnames(tickets)
#Examine structure of the data
str(tickets)
```

##Summary of the Dataset 

The dataset consists of San Francisco Municipal Transportation Agency (SFMTA) driver reported transit only lane violations from March 2008 to February 2015. The dataset contains `r NROW(tickets)` observations and `r NCOL(tickets)` variables. The data were obtained from the US government's open data site at https://catalog.data.gov/dataset/muni-driver-reported-transit-only-lane-violations. 

###Description of Variables in the dataset: 

* `Object.ID` -- Unique key that corresponds to each observation
* `Ticket.Number` -- Unique number that corresponds to each ticket catalogued with the Superior Court of San Francisco
* `Citation.Issue.Date` -- The date (MM/DD/YY) that the traffic only lane violation citation was issued
* `Citation.Issue.Month` -- The month (MM) that the traffic only lane violation citation was issued; This column was completely blank when downloaded
* `Citation.Issue.Time` -- The time of day (HH:MM) PST that the traffic only lane violation citation was issued
* `Location` -- The street address where the traffic only lane violation citation was issued
* `Violation.Code` -- A code that corresponds to the type of traffic only lane violation that was committed to warrant issuing a citation
* `Violation` -- Description of the traffic only lane violation issued in the citation
* `Fine.Amount` -- The dollar amount of to be paid for the traffic only lane violation
* `Citation.Status` -- Status of the citation (Open, Closed, Unapplied)
* `Amount.Paid` -- The dollar amount paid for the traffic only lane violation
* `Amount.Due` -- Outstanding balance for the traffic only lane violation 
* `Suspend.Code` -- Code that represents reason for license suspension (Suspended until the driver goes to driving school, pays fine, too many points on license, etc.)
* `Suspend.Process.Date` -- Date the license suspension goes into effect
* `Suspend.Until.Date` -- Date the license suspension is over
* `Disposition.Code` -- The city of San Francisco does not release disposition codes
* `Last.Edited.Date` -- Date that represents the last date the corresponding observation was edited; This column was completely blank when downloaded
* `Geom` -- The latitude and longitude (latitutde, longitude) of the traffic only lane violation 



###Explanations of Violation variable:
 * 'DBL Park' -- Double Parking
 * 'TWAWY ZN#1' and 'TWAWY ZN#2' -- Towaway Zone 1 and Towaway Zone 2
 * 'TRNST ONLY' -- Parking Transit-Only
 * 'PRK PROHIB' -- Tow-Away Zone - Downtown Core
 * 'PK PHB OTD' -- Tow-Away Zone - Outside Downtown Core
 * 'PK FR LN' -- Parking in Fire Lane
 * 'ON SIDEWLK' -- Parking on Sidewalk
 * 'DBL PARK' -- Double Parking
 * 'BUS ZONE' -- Parking in Bus Zone
 * 'OVR 18" C' -- Over 18 inches from Curb
 
 
## Formatting Data
```{r Reformatting Variables, tidy=T, results='hide', message=FALSE}
# Format Date and Month columns
#Change citation issue date from string to month/day/year date variable 
tickets$Citation.Issue.Date <- mdy(tickets$Citation.Issue.Date)
#Populate blank column for citation issue month with month of citation issue date
tickets$Citaton.Issue.Month <- NULL
tickets$Citation.Issue.Month <- month(tickets$Citation.Issue.Date, label = TRUE)

# Convert Amount.Paid from factor to numeric
tickets$Amount.Paid <- as.numeric(tickets$Amount.Paid)
```


## Data Cleaning
```{r Data Cleaning, tidy=T}
# New Column "Paid.On.Time" (1 = true, 0 = false). Based on assumption that late payments have to pay an extra fine and, thus, Amount.Paid > Fine.Ammount
tickets$Paid.On.Time <- ifelse(tickets$Amount.Paid > tickets$Amount.Due, 1, 0)
tickets$Paid.On.Time <- as.factor(tickets$Paid.On.Time)

# Seperate Geom variable into 2 separate latitude and longitude numeric variables 
  ## remove parentheses from Geom variable
tickets$Geom <- substr(tickets$Geom, 2, nchar(as.character(tickets$Geom))-1)
  ## split string at comma & create new columns
out <- strsplit(as.character(tickets$Geom), ", ")
  ##Drop observations with no Geom data. ***21 OBSERVATIONS DROPPED***
tickets <- tickets[!(is.na(tickets$Geom) | tickets$Geom == ""), ]
tickets <- data.frame(tickets, do.call(rbind, out))
rm(out)
  ## rename columns and set str to numeric
names(tickets)[names(tickets) == "X1"] <- "Latitude"
names(tickets)[names(tickets) == "X2"] <- "Longitude"
tickets$Latitude <- as.numeric(as.character(tickets$Latitude))
tickets$Longitude <- as.numeric(as.character(tickets$Longitude))
head(tickets)

# Create column indicating the year the ticket was issued
tickets$Citation.Issue.Year <- year(tickets$Citation.Issue.Date)

# Create column indicating DayOfWeek (Sunday (1) to Saturday (7)) that the ticket was issued
tickets$Citation.Issue.DayOfWeek <- wday(tickets$Citation.Issue.Date)

# Create column indicating whether the ticket was issued on: weekday (1), weekend (0)
tickets$Citation.Issue.Weekday <- ifelse(tickets$Citation.Issue.DayOfWeek >= 2 & tickets$Citation.Issue.DayOfWeek <= 6, 1, 0)
tickets$Citation.Issue.Weekday <- as.factor(tickets$Citation.Issue.Weekday)
  #Code levels of Weekday variable
levels(tickets$Citation.Issue.Weekday) <- list("Weekend" = 0 , "Weekday" = 1)

# Convert Citation.Issue.DayOfWeek from 1, 2, 3, 4, 5, 6, 7 to Mon, Tues, Wed, Thurs, Fri, Sat
tickets$Citation.Issue.DayOfWeek <- wday(tickets$Citation.Issue.Date, label = TRUE)

# Create column for Hour of citation issuance. Drop observations with no Citation.Issue.Time data. ***0 OBSERVATIONS DROPPED***
tickets <- tickets[!(is.na(tickets$Citation.Issue.Time) | tickets$Citation.Issue.Time == ""), ]
outTime <- strsplit(as.character(tickets$Citation.Issue.Time), ":")
tickets <- data.frame(tickets, do.call(rbind, outTime))
names(tickets)[names(tickets) == "X1"] <- "Citation.Issue.Hour"
names(tickets)[names(tickets) == "X2"] <- "Citation.Issue.Minute"
tickets$Citation.Issue.Hour <- as.numeric(as.character(tickets$Citation.Issue.Hour))
tickets$Citation.Issue.Minute <- NULL
rm(outTime)

# Create column from Month-Year
tickets <- within(tickets, Citation.Issue.MonthYear <- paste(Citation.Issue.Month, "-", Citation.Issue.Year, sep=""))

# Drop "Unapplied" and empty/NA observations within Citation.Status. Unapplied result is outside the scope of this project. ***84 OBSERVATIONS DROPPED***
tickets <- tickets[!(is.na(tickets$Citation.Status) | tickets$Citation.Status =="" | tickets$Citation.Status == "Unapplied"),]
levels(tickets$Citation.Status) <- list("Open" = "Open", "Closed" = "Closed")

# Drop extraneous variables: Object.ID (just row index), Location (Not need street address b/c we have latitude and longitude), Last.Edited.Date (empty column), Suspend.Code (NA), Disposition.Code (NA), Violation.Code (have this info in Violation variable), Citation.Issue.Time (Not needed for analysis), Suspend.Process.Date and Suspend.Until.Date (outside scope of analysis), Geom (we split this into latitude and longitude and have all this info already)
tickets$Object.ID <- NULL
tickets$Location <- NULL
tickets$Last.Edited.Date <- NULL
tickets$Suspend.Code <- NULL
tickets$Disposition.Code <- NULL
tickets$Violation.Code <- NULL
tickets$Citation.Issue.Time <- NULL
tickets$Suspend.Process.Date <- NULL
tickets$Suspend.Until.Date <- NULL
tickets$Geom <- NULL

#Violation Variable 
table(tickets$Violation)
  # Drop "NO VIOL" in Violation variable. ***1 OBSERVATION DROPPED***
  # Drop "ON SIDEWLK" in Violation variable. ***1 OBSERVATION DROPPED***
  # Drop "OVR 18" C" in Violation variable. ***5 OBSERVATIONS DROPPED***
  # Drop "PK FR LN" in Violation variable. ***1 OBSERVATION DROPPED***
  # Drop "TRNST ONLY" in Violation variable. ***70 OBSERVATION DROPPED***
tickets <- tickets[!table(tickets$Violation)[tickets$Violation] <= 70,]
table(tickets$Violation)
  #Recode factor varibale without the Null categories
levels(tickets$Violation) <- list("BUS ZONE" = "BUS ZONE", "DBL PARK" = "DBL PARK", "PK PHB OTD" = "PK PHB OTD", "PRK PROHIB" = "PRK PROHIB", "TWAWY ZN#1" = "TWAWY ZN#1", "TWAWY ZONE" = "TWAWY ZONE")


# Create neighborhood column using Lat/Long and Census Block Data from shp file
## Thanks to Shekeine for instructions: http://stackoverflow.com/questions/29872109/binning-longitude-latitude-labeled-data-by-census-block-id
## Load Census tract data for San Fransisco
#tractSF <- shapefile('Analysis Neighborhoods - 2010 census tracts assigned to neighborhoods/geo_export_72577790-b6d5-439f-9b51-edc69b942ac6.shp')
```

###Dropped Observations

* We dropped all observations with no Geom data
    * This was only 21 of the 17,178 observations
* We dropped all observations with no information on the time of the citation
    * Every observation had information on the time of the citation
* We dropped all observations where citation status was not specified or "unapplied"
    * We determined "Unapplied" citations are outside the scope of this project
    * This only led to 84 observations being dropped
* Dropping extraneous variables did not affect the number of observations in the dataset
* We dropped Violations where the listed violation was determined to be "No Violation" or occured less than or equal to 70 times
    * This led to 78 observation being dropped
    * It is worth noting that only 1 violation was deemed not to be a violation

Overall, we dropped `r 78 + 84 + 21` of the original 17,178 observations in the dataset. This only accounts for `r round(((78 + 84 + 21)/17178)*100,2)` percent of the dataset. Thus, we are confident that we did not introduce any inherent biases to the data by dropping the aforementioned observations. 

##Cleaned Dataset

```{r Cleaned Dataset Properties, tidy=T, results='hide', message=FALSE}
#Check dimensions of cleaned dataset 
dim(tickets)
#Check all variable names
colnames(tickets)
#Examine structure of cleaned dataset
str(tickets)
```


After cleaning the dataset to begin exploratory data anlysis, there are `r NROW(tickets)` observations and `r NCOL(tickets)` variables.

# Mike EDA 

```{r EDA of Time variables, tidy=T, message=F}

# Month
table(tickets$Citation.Issue.Month)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Month))
  ##Without 2015 
tickets_to2014 <- filter(tickets, tickets$Citation.Issue.Year < 2015)
table(tickets_to2014$Citation.Issue.Month)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Month))
  ##Without 2008 and 2015
tickets_no_partials <- filter(tickets, tickets$Citation.Issue.Year > 2008 & tickets$Citation.Issue.Year < 2015)
table(tickets_no_partials$Citation.Issue.Month)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Month))

# Year
table(tickets$Citation.Issue.Year)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Year))
#  *****There are only 37 violations for the beginning of 2015. Should be dropped*****
#  *****Could also consider dropping 2008 as that is a partial year as well. (157)****** 
  ##Without 2015
table(tickets_to2014$Citation.Issue.Year)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Year))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Year)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Year))

# Day of Week 
table(tickets$Citation.Issue.DayOfWeek)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.DayOfWeek))
  ##Without 2015
table(tickets_to2014$Citation.Issue.DayOfWeek)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.DayOfWeek))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.DayOfWeek)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.DayOfWeek))
# ******* Let's look at how we coded the Day of Week variable. This trend shoud be lower on Sunday ******

# Weekday
table(tickets$Citation.Issue.Weekday)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Weekday))
  ##Without 2015
table(tickets_to2014$Citation.Issue.Weekday)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Weekday))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Weekday)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Weekday))
# ******* Let's look at how we coded the Day of Week variable. This trend shoud be lower on Sunday ******

#Hour 
table(tickets$Citation.Issue.Hour)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Hour))
  ##Without 2015
table(tickets_to2014$Citation.Issue.Hour)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Hour))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Hour)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Hour))
# People are getting boned during rush hour on the way home

#Summary stats for the different tickets datasets for years
summary(tickets)
summary(tickets_to2014)
summary(tickets_no_partials)
```

```{r EDA of Money and Violation variables, tidy=T, message=F}
#Violations
table(tickets$Violation)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation))
  #Violations broken down by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge")
  ##Without 2015
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Violation, fill = as.factor(tickets_to2014$Citation.Issue.Year)), position = "dodge")
  ##Without 2008 and 2015
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Violation, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "dodge")

#Fine amount
table(tickets$Fine.Amount)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Fine.Amount))
  ##By Violation (without 2008 and 2015)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount)) + facet_wrap(~as.factor(tickets_no_partials$Violation))
  ## It is clear that tickets have a low fine amount and high fine amount. Is it time of day, weekday vs weekend, year, etc.?
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "fill") + facet_wrap(~tickets_no_partials$Violation)

ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount, fill = tickets_no_partials$Citation.Issue.Weekday), position = "fill") + facet_wrap(~tickets_no_partials$Violation)


#Citation Status
table(tickets_no_partials$Citation.Status)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Status))

#Amount paid
table(tickets$Amount.Paid)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Amount.Paid))
  #Why are some amounts paid so low? Are these cases where status is open still? 
tickets_by_paid_status <- tickets %>% group_by(Amount.Paid, Citation.Status) 
summary(tickets_by_paid_status)
ggplot(data = tickets_by_paid_status) + geom_bar(mapping = aes(x = tickets$Amount.Paid, fill = tickets$Citation.Status))
  #Most of these ones are still open but it still doesn't make sense
  #What types of violations are these????
tickets_paid_viol <- tickets %>% group_by(Amount.Paid, Violation)
ggplot(data = tickets_paid_viol) + geom_bar(mapping = aes(x = tickets$Amount.Paid, fill = tickets$Violation))
  #Looks like prices of violations vary. By time? Weekend? 

#Amount Due
table(tickets$Amount.Due)
ggplot(data = tickets) + geom_histogram(mapping = aes(x = tickets$Amount.Due, fill = tickets$Violation))
  ##The vast majority of tickets are fully paid off
head(table(tickets$Amount.Due)) 
  ##By citations status?
tickets_due_status <- tickets %>% group_by(Amount.Due, Citation.Status)
summary(tickets_due_status)
ggplot(data = tickets_due_status) + geom_histogram(mapping = aes(x = tickets$Amount.Due, fill = tickets$Citation.Status))
  #That seems to tell the whole story


#Paid on time
table(tickets$Paid.On.Time)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Paid.On.Time))



```


##How have the prices of TOLE violations changed over time?

```{r prices change over time EDA, tidy=T, message=F}
#summary stats and bar chart of fine amount
summary(tickets$Violation)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation))

  #Violations broken down by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge")
  #No partial years
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Violation, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "dodge")

#Because the distribution of fine amount has several high-end outliers, we should be using the median as our measure of central tendency. This is even more clear when comparing the difference between the mean and median (112.7 and 85). 
median(tickets$Fine.Amount)
mean(tickets$Fine.Amount)
median(tickets_no_partials$Fine.Amount)
mean(tickets_no_partials$Fine.Amount)

# find the median fine amount by violation
fine_by_viol <- tickets %>% 
  group_by(Violation) %>% 
  summarise(median = median(Fine.Amount))
fine_by_viol

#Maybe put these results in a table?

# find the median fine amount by year
fine_by_year <- tickets %>%
  group_by(Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year)

#plot median fine amount by year to see trend in prices over time
fine.plot.year <- ggplot(data = fine_by_year, aes(x = Citation.Issue.Year, y = median))
fine.plot.year <- fine.plot.year + geom_line() + geom_point() 
fine.plot.year
# *******Median fine amount per year seems to have increased significantly over time. Index with inflation to see if this is really the case?*******

#fine amount by year, violation
fine_by_year_viol <- tickets %>% 
  group_by(Violation,Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year_viol)

#plot fine by year and violation to see change and trends
fine.plot.yv <- ggplot(data = fine_by_year_viol, aes(x = Citation.Issue.Year, y = median))
fine.plot.yv <- fine.plot.yv + geom_line() + geom_point() 
fine.plot.yv <- fine.plot.yv + facet_wrap(~ Violation)
fine.plot.yv

# ********** Looks like we should look at violations and the years they occur in and maybe get more specific on type of violation*********

```


##Are more citations issued on certain days of the week? Are more tickets issued on weekdays than weekends?

