---
title: "Project_1_EDA"
author: "Michael Arango, Yeshwant Chillakuru, Jackson Crum"
date: "March 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

----------------------------------------------------------------------

# Yesh's Stuff
## What traffic only lane violations are most common?
```{r violations, tidy=T, message=FALSE, echo=FALSE}

tickets <- read.csv("tickets.csv", header=TRUE)

ggplot(data=tickets, aes(Violation)) + geom_bar() + coord_flip() + geom_text(stat="count", aes(label=..count..))
```

## What traffic lane violations make up the most revenue?
```{r violations2, tidy=T, message=FALSE, echo=FALSE}


tickets.violation <- aggregate(tickets[,c("Amount.Paid")], by=list(tickets$Violation), FUN=sum)
tickets.violation <- rename(tickets.violation, Violation = Group.1, Total.Revenue = x)

ggplot(tickets.violation, aes(x=Violation, y=Total.Revenue)) + geom_bar(stat="identity") + coord_flip() + geom_text(aes(label=Total.Revenue))

rm(tickets.violation)

```

## What violations produce the most daily revenue on weekends and weekdays?
```{r violations3, tidy=T, message=FALSE, echo=FALSE}

tickets.violWeekday <- aggregate(tickets[,c("Amount.Paid")], by=list(tickets$Violation, tickets$Citation.Issue.Weekday), FUN=sum)
tickets.violWeekday <- rename(tickets.violWeekday, Violation = Group.1, Weekday = Group.2, Total.Revenue = x)
tickets.violWeekday$Daily.Average.Revenue <- ifelse(tickets.violWeekday$Weekday == 1, tickets.violWeekday$Total.Revenue/5, tickets.violWeekday$Total.Revenue/2)
tickets.violWeekday$Weekday <- ifelse(tickets.violWeekday$Weekday == 0, "Weekend", "Weekday")

p <- ggplot(tickets.violWeekday, aes(x=Violation, y=Daily.Average.Revenue, fill=Weekday))
p + geom_bar(stat="identity") + coord_flip()

rm(p, tickets.violWeekday)
```

## What times get the most tickets for the top 3 violation types?
```{r violations4, tidy=T, message=FALSE, echo=FALSE}

ViolCount <- tickets %>% group_by(Violation) %>% summarise(freq=n())
topViol <- ViolCount[order(-ViolCount$freq),]
topViol <- topViol[1:3,]


# tickets.topViol <- subset(tickets, tickets$Violation == topViol[1,1] | tickets$Violation == topViol[2,1] | tickets$Violation == topViol[3,1])
tickets.topViol <- subset(tickets, tickets$Violation == topViol$Violation[1] | tickets$Violation == topViol$Violation[2] | tickets$Violation == topViol$Violation[3])

tickets.topViol <- rename(tickets.topViol, Top.3.Violations = Violation)

p1 <- ggplot(tickets.topViol, aes(Citation.Issue.Hour, fill=Top.3.Violations))
p1 + geom_density(alpha = .6)

```

## Which fees get the most tickets for the top 3 violation types?
```{r violations5, tidy=T, message=FALSE, echo=FALSE}


p2 <- ggplot(tickets.topViol, aes(Amount.Paid, fill=Top.3.Violations))
p2 + geom_density(alpha = .6)

rm(p1, p2, topViol, ViolCount, tickets.topViol)
```

# Mike EDA 

```{r EDA of Time variables, tidy=T, message=F}

# Month
table(tickets$Citation.Issue.Month)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Month))
  ##Without 2015 
tickets_to2014 <- filter(tickets, tickets$Citation.Issue.Year < 2015)
table(tickets_to2014$Citation.Issue.Month)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Month))
  ##Without 2008 and 2015
tickets_no_partials <- filter(tickets, tickets$Citation.Issue.Year > 2008 & tickets$Citation.Issue.Year < 2015)
table(tickets_no_partials$Citation.Issue.Month)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Month))

# Year
table(tickets$Citation.Issue.Year)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Year))
#  *****There are only 37 violations for the beginning of 2015. Should be dropped*****
#  *****Could also consider dropping 2008 as that is a partial year as well. (157)****** 
  ##Without 2015
table(tickets_to2014$Citation.Issue.Year)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Year))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Year)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Year))

# Day of Week 
table(tickets$Citation.Issue.DayOfWeek)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.DayOfWeek))
  ##Without 2015
table(tickets_to2014$Citation.Issue.DayOfWeek)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.DayOfWeek))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.DayOfWeek)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.DayOfWeek))
# ******* Let's look at how we coded the Day of Week variable. This trend shoud be lower on Sunday ******

# Weekday
table(tickets$Citation.Issue.Weekday)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Weekday))
  ##Without 2015
table(tickets_to2014$Citation.Issue.Weekday)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Weekday))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Weekday)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Weekday))
# ******* Let's look at how we coded the Day of Week variable. This trend shoud be lower on Sunday ******

#Hour 
table(tickets$Citation.Issue.Hour)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Hour))
  ##Without 2015
table(tickets_to2014$Citation.Issue.Hour)
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Citation.Issue.Hour))
  ##Without 2008 and 2015
table(tickets_no_partials$Citation.Issue.Hour)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Issue.Hour))
# People are getting boned during rush hour on the way home

#Summary stats for the different tickets datasets for years
summary(tickets)
summary(tickets_to2014)
summary(tickets_no_partials)
```

```{r EDA of Money and Violation variables, tidy=T, message=F}
#Violations
table(tickets$Violation)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation))
  #Violations broken down by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge")
  ##Without 2015
ggplot(data = tickets_to2014) + geom_bar(mapping = aes(x = tickets_to2014$Violation, fill = as.factor(tickets_to2014$Citation.Issue.Year)), position = "dodge")
  ##Without 2008 and 2015
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Violation, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "dodge")

#Fine amount
table(tickets$Fine.Amount)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Fine.Amount))
  ##By Violation (without 2008 and 2015)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount)) + facet_wrap(~ tickets_no_partials$Violation)
  ## It is clear that tickets have a low fine amount and high fine amount. Is it time of day, weekday vs weekend, year, etc.?
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "fill") + facet_wrap(~tickets_no_partials$Violation)

ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Fine.Amount, fill = tickets_no_partials$Citation.Issue.Weekday), position = "fill") + facet_wrap(~tickets_no_partials$Violation)


#Citation Status
table(tickets_no_partials$Citation.Status)
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Citation.Status))

#Amount paid
table(tickets$Amount.Paid)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Amount.Paid))
  #Why are some amounts paid so low? Are these cases where status is open still? 
tickets_by_paid_status <- tickets %>% group_by(Amount.Paid, Citation.Status) 
summary(tickets_by_paid_status)
ggplot(data = tickets_by_paid_status) + geom_bar(mapping = aes(x = tickets$Amount.Paid, fill = tickets$Citation.Status))
  #Most of these ones are still open but it still doesn't make sense
  #What types of violations are these????
tickets_paid_viol <- tickets %>% group_by(Amount.Paid, Violation)
ggplot(data = tickets_paid_viol) + geom_bar(mapping = aes(x = tickets$Amount.Paid, fill = tickets$Violation))
  #Looks like prices of violations vary. By time? Weekend? 

#Amount Due
table(tickets$Amount.Due)
ggplot(data = tickets) + geom_histogram(mapping = aes(x = tickets$Amount.Due, fill = tickets$Violation))
  ##The vast majority of tickets are fully paid off
head(table(tickets$Amount.Due)) 
  ##By citations status?
tickets_due_status <- tickets %>% group_by(Amount.Due, Citation.Status)
summary(tickets_due_status)
ggplot(data = tickets_due_status) + geom_histogram(mapping = aes(x = tickets$Amount.Due, fill = tickets$Citation.Status))
  #That seems to tell the whole story


#Paid on time
table(tickets$Paid.On.Time)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Paid.On.Time))



```


##How have the prices of TOLE violations changed over time?

```{r prices change over time EDA, tidy=T, message=F}
#summary stats and bar chart of fine amount
summary(tickets$Violation)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation))

  #Violations broken down by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge")
  #No partial years
ggplot(data = tickets_no_partials) + geom_bar(mapping = aes(x = tickets_no_partials$Violation, fill = as.factor(tickets_no_partials$Citation.Issue.Year)), position = "dodge")

#Because the distribution of fine amount has several high-end outliers, we should be using the median as our measure of central tendency. This is even more clear when comparing the difference between the mean and median (112.7 and 85). 
median(tickets$Fine.Amount)
mean(tickets$Fine.Amount)
median(tickets_no_partials$Fine.Amount)
mean(tickets_no_partials$Fine.Amount)

# find the median fine amount by violation
fine_by_viol <- tickets %>% 
  group_by(Violation) %>% 
  summarise(median = median(Fine.Amount))
fine_by_viol

#Maybe put these results in a table?

# find the median fine amount by year
fine_by_year <- tickets %>%
  group_by(Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year)

#plot median fine amount by year to see trend in prices over time
fine.plot.year <- ggplot(data = fine_by_year, aes(x = Citation.Issue.Year, y = median))
fine.plot.year <- fine.plot.year + geom_line() + geom_point() 
fine.plot.year
# *******Median fine amount per year seems to have increased significantly over time. Index with inflation to see if this is really the case?*******

#fine amount by year, violation
fine_by_year_viol <- tickets %>% 
  group_by(Violation,Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year_viol)

#plot fine by year and violation to see change and trends
fine.plot.yv <- ggplot(data = fine_by_year_viol, aes(x = Citation.Issue.Year, y = median))
fine.plot.yv <- fine.plot.yv + geom_line() + geom_point() 
fine.plot.yv <- fine.plot.yv + facet_wrap(~ Violation)
fine.plot.yv

# ********** Looks like we should look at violations and the years they occur in and maybe get more specific on type of violation*********

```


##Are more citations issued on certain days of the week? Are more tickets issued on weekdays than weekends?

