---
title: "Analysis of the San Fransico TOLE Program"
author: "Michael Arango, Yeshwant Chillakuru, Jackson Crum"
date: "March 6, 2017"
output:
  pdf_document: default
---

```{r loadData, include =FALSE}
knitr::opts_chunk$set(fig.pos = 'H')

library(ggplot2)
library(dplyr)
library(scales)

rawTickets <- data.frame(read.csv('Muni_Driver_Reported_Transit_Only_Lane_Violations.csv', header = T, sep = ","))
tickets <- read.csv("tickets.csv")

```


# Background Information on Dataset

## General Information and Data Collection Process
The Transit-Only Lane Enforcement (TOLE), run by the San Fransico Municipal Transportation Agency (SFMTA) from March 2008 to February 2015, is a pilot program to reduce obstruction of transit-only lanes through increased enforcement. Currently, private and commerical vehicles that drive and park in transit-only lanes can cause delays in bus service, increases in traffic congestion, and inconveniences to customers.

Under the TOLE program, all public buses are equipped with front-facing video cameras to capture the license plates of vehicles parked or stopped in a dedicated transit-lane. The vehicle's registered owner will then recieve the citation in the mail as a means to deter future transit-only lane violations. These citations and their accompanying details form the dataset we are examining.[^1]

[^1]:Data retrieved from Data.gov (<https://catalog.data.gov/dataset/muni-driver-reported-transit-only-lane-violations>) 

San Fransico currently operates 26 miles of transit-only lanes with 160,000 daily passsengers and plans future expansion of 22 miles, so the TOLE program has strong potential for long-term impact by increasing the efficiency of the municipal transit system and potentially increasing municipal revenue temporarily as drivers adjust to the enforcement scheme.[^2]

[^2]:San Francisco Municipal Transportation Agency, "Transit Only Lane Enforcement," SFMTA, accessed on February 24, 2017, <https://www.sfmta.com/sites/default/files/projects/2015/TOLE%20one%20pager_v3.pdf> 

## Description of Dataset

The unit of observation was an individual citation for a transit-only lane violation reported by municipal drivers under the TOLE program, and the raw dataset contains the following variables regarding the transit-only lane violations:

* `Object.ID` -- Unique key that corresponds to each observation
* `Ticket.Number` -- Unique number that corresponds to each ticket cataloged with the Superior Court of San Francisco
* `Citation.Issue.Date` -- The date (MM/DD/YY) that the traffic only lane violation citation was issued
* `Citation.Issue.Month` -- The month (MM) that the traffic only lane violation citation was issued; This column was completely blank when downloaded
* `Citation.Issue.Time` -- The time of day (HH:MM) PST that the traffic only lane violation citation was issued
* `Location` -- The street address where the traffic only lane violation citation was issued
* `Violation.Code` -- A code that corresponds to the type of traffic only lane violation that was committed to warrant issuing a citation
* `Violations` -- Description of the traffic only lane violation issued in the citation
* `Fine.Amount` -- The dollar amount of to be paid for the traffic only lane violation. Mean: `r round(mean(tickets$Fine.Amount),2)` USD.
* `Citation.Status` -- Status of the citation (Open, Closed, Unapplied)
* `Amount.Paid` -- The dollar amount paid for the traffic only lane violation. Mean: `r round(mean(tickets$Amount.Paid),2)` USD.
* `Amount.Due` -- Outstanding balance for the traffic only lane violation. Mean: `r round(mean(tickets$Amount.Due),2)` USD.
* `Suspend.Code` -- Code that represents reason for license suspension (Suspended until the driver goes to driving school, pays fine, too many points on license, etc.)
* `Suspend.Process.Date` -- Date the license suspension goes into effect
* `Suspend.Until.Date` -- Date the license suspension is over
* `Disposition.Code` -- The city of San Francisco does not release disposition codes
* `Last.Edited.Date` -- Date that represents the last date the corresponding observation was edited; This column was completely blank when downloaded
* `Geom` -- The latitude and longitude (latitude, longitude) of the traffic only lane violation 

## Limitations of Dataset and Data Cleaning

The limitations of the data include the timing of the implementation of the video camera citation system. While SFMTA claims 100% of their buses are equipped with cameras, they fail to indicate if the 100% implemention was at the start or end of the program. We believe there was a trial period at the beginning, so we decided to drop variables in 2008 when the program first started because there were only 154 citations issued that year, but typically several thousand citations issued in the following years. In addition our dataset lacked demographic controls for the regions where citations were issued. This means demographic factors could cause unobserved biases in our data. Similarly to the domestic controls, the dataset lacks observations of citations issued before the program was put in place, which prevents a before and after analysis. While out of the scope of this initial analysis, future analysis can map to the `Geom` variable in the dataset to census tracts to derive demographic information and could be used to compare to other citations.

Some observations were missing key data. We decided to drop all observations with no `Geom` data, so that future geospatial and demographic analysis could be done (12 observations). We dropped all observations where `Citation.Status` was not specificied or "unapplied", as we determined "unapplied" was out of the scope of this project (84 observations). We dropped `Violations` where the listed violation was determined to be "No Violation" (1 observation) or occured less than or equal to 70 times (77 observations). We also dropped citations in the year 2015 to avoid biases in monthly and yearly analysis because only 2 months were included for 2015.

Overall, we dropped `r 191 + 78 + 84 + 21` of the original 17,178 observations in the dataset. This only accounts for `r round(((191 + 78 + 84 + 21)/17178)*100,2)` percent of the dataset. Thus, we are confident that we did not introduce any inherent biases to the data by dropping the aforementioned observations. 

## Background Prior External Analysis Related to Content of Dataset

Most of the analysis compared TOLE data to traffic flow rates. According to SFMTA, certain corridors experienced reductions in delays after TOLE implemention. Westbound Geary Street corridor saw a 3% daily and 7% afternoon reduction in delays. Westbound Stutter Street saw a 15% daily and 20% afternoon reduction in delays. However, this analysis lacks significant meaning because it cherry picks certain areas and also lacks a control comparison for other areas that may not have been in the TOLE program. Additionally, their analysis has some clear flaws as they state that of the 1700 different vehicles that recieved a citation from March to August of 2012, only 2% recieved a citation during the "same three-month period in 2013." Again this analysis lacks a control comparison and suffers a clear inconsistancy as March to August is a six-month period---not three month as SFMTA stated.[^2]

According to Michael Rhodes, San Francisco's transit-only lane violations are generally unenforced. While SFMTA had not released citation data prior to Rhodes' article in 2009, he cites anecdotal evidence from interviews with municipal drivers and San Francisco Police Department. Based on the municipality's weak analysis and the anecdotal evidence provided by Rhodes of the need for an effective transit-only lane violation enforcement scheme, a further analysis of municipality data must be conducted. While we cannot analyze a before and after comparison of the TOLE program on transit-only lane violations or traffic flow improvement, we can examine the distribution of violations by violation type, time of day, day of the week, and revenue potential.


Source2: http://sf.streetsblog.org/2009/08/11/violations-in-sfs-transit-only-lanes-rampant-and-rarely-enforced/

```{r citation2}
print("Cite the above sources properly")
```
  
## Broad Questions

The TOLE program was implemented in 2008 as a pilot set to expire in 2012, unless extended upon an evaluation in 2011. In 2011, the City of San Francisco decided to vote to implement the TOLE program permanently. Therefore, understanding the effects and dimensions of the program are vital to constantly improving it. We believed in an exploratory analysis that looking at revenue generation would be in ideal means of understanding this program. Revenue optimization both increases funding for the state and creates economic incentive for drivers' to avoid citations. Under properly placed incentives and fine structures, revenue optimization can lead to ideal traffic outcomes.

_**Broad Question:** Under what paramters does the TOLE program optimize TOLE program revenues?_

We started with this broad question before choosing a full SMART (Specific, Measurable, Answerable, Relevant, Time bound) question to allow EDA to inform the development of the SMART question.

Source: http://www.leginfo.ca.gov/pub/11-12/bill/asm/ab_1001-1050/ab_1041_cfa_20110429_113836_asm_floor.html

```{r citation3}
print("Cite the above sources properly")
```

\newpage
## EDA & SMART Question

We first examined basic summary statistics involving our dataset. Our cleaned dataset had `r dim(tickets)[1]` observations (i.e. citations) and `r dim(tickets)[2]` variables. There were 5 violation:
 * 'BUS ZONE' -- Parking in Bus Zone  -- `r table(tickets$Violation)["BUS ZONE"]` citations
 * 'DBL Park' -- Double Parking -- `r table(tickets$Violation)["DBL PARK"]` citations
 * 'PK PHB OTD' -- Tow-Away Zone - Outside Downtown Core  -- `r table(tickets$Violation)["PK PHB OTD"]` citations
 * 'PRK PROHIB' -- Tow-Away Zone - Downtown Core  -- `r table(tickets$Violation)["PRK PROHIB"]` citations
 * 'TWAWY ZN#1' -- Towaway Zone 1 and Towaway Zone  -- `r table(tickets$Violation)["TWAWY ZN#1"]` citations
 * 'TWAWY ZONE' -- Towaway Zone  -- `r table(tickets$Violation)["TWAWY ZONE"]` citations

We then examined if violations vary by year, and saw that `PRK PROHIB` and `PK PHB OTD` occur only in after 2013, while `TWAWY ZN#1` and `TWAWY ZONE` violations were issued only until 2013. 

```{r EDA1, echo=FALSE}
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge", color="black") + coord_flip() + xlab("Violation") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Number of Tickets by Violation and Year of Citation") + labs(fill = "Citation Issue Year") 
```

\newpage
We decided to examine how fine amount varies within each fine and discovered a bimodal distribution with a "high" and "low" fine.

```{r EDA2, echo=FALSE, message=FALSE, tidy=TRUE}
fine.plot <- ggplot(data = tickets) + geom_histogram(mapping = aes(x = tickets$Fine.Amount), fill="salmon")

fine.plot + facet_wrap(~tickets$Violation) + xlab("Fine Amount") + theme(plot.title = element_text(hjust = 0.5)) + 
    ggtitle("Histogram of Fine Amount by Violation")

rm(fine.plot)
```

\newpage
However, upon further examination, we saw there was no obvious difference in `Fine.Amount` when controlling for other factors like time of day:

```{r EDA3, echo=FALSE, message=FALSE, tidy=TRUE}
# Initialize new factor variable Fine.Level
tickets$Fine.Level <- tickets$Fine.Amount
tickets$Fine.Level <- ifelse(tickets$Fine.Amount < 175, "Low", "High")
tickets$Fine.Level <- as.factor(tickets$Fine.Level)
tickets$Fine.Level <- relevel(tickets$Fine.Level, "Low")
## Generate table

fine.plot3 <- ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Hour), 
    fill = "salmon", color="black")
fine.plot3 + facet_wrap(~tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + 
    ggtitle("Bar Chart of Fine Level by Hour of Citation")

rm(fine.plot3)
```

\newpage
We then began to shift our analysis towards looking at the interaction between time and revenue generated by fines. We looked at the distribution of different citations over time in a stacked bar plot:


```{r EDA4, echo=FALSE, message=FALSE}

ggplot(tickets, aes(x = Citation.Issue.Hour, fill = Violation)) + geom_bar(position = "fill") + 
    scale_x_continuous(breaks = seq(0, 24, 1)) + xlab("Hour of Citation") + 
    theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Stacked Bar Chart of Violations by Citation Issue Hour ")

```

This graph shows that Double Parking (`DBL PARK`) and Bus Zone (`BUS ZONE`) Violations comprimise most non-rush hour violations, while Tomaway Zone (`TWAWY ZN#1` and `TWAWY ZONE`) and Downtown Towaway Zone (`PRK PROHIB`) Violations make up most of rush hour violations.

\newpage
We then investigated which violations produce the most revenue. On a daily basis, Towaway Zone 1 produces the most revenue. even when broken down to weekdays and weekends.

```{r EDA5, echo=FALSE}
#Create variable that stores sum of apount paid by violation and weekday 
tickets.violWeekday <- aggregate(tickets[,c("Amount.Paid")], by=list(tickets$Violation, tickets$Citation.Issue.Weekday), FUN=sum)
  ##Rename the columns of tickets.violWeekday
tickets.violWeekday <- rename(tickets.violWeekday, Violation = Group.1, Weekday = Group.2, Total.Revenue = x)
  ##Create new column for daily average revenue: IF day == weekday, then divide TR by 5; ELSE divide by 2
tickets.violWeekday$Daily.Average.Revenue <- ifelse(tickets.violWeekday$Weekday == 1, tickets.violWeekday$Total.Revenue/5, tickets.violWeekday$Total.Revenue/2)

#Average Weekday and Weekend Revenue by Violation
p <- ggplot(tickets.violWeekday, aes(x=Violation, y=Daily.Average.Revenue, fill=Weekday))
p + geom_col() + coord_flip() + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Total Average Weekday and Weekend Revenue by Violation") + ylab("Daily Average Revenue")

#Remove from Workspace
rm(p, tickets.violWeekday)
```

We saw that there was a potential for certain days of the week to generate more revenue than others so we decided to focus on how revenue can vary based on each day of the week. Understanding the deviation of mean revenue for weekday from the mean daily revenue is an ideal starting point for further analysis. If deviation is present, then future work could examine variation based on violation type, time of day, rush hour vs non-rush hour. Additionally, using median is justified to account for the biasing effect of the uneven bimodal distribution of fine amounts (the high and low fines) and outher outliers. Ideally, if the dataset could be connected to traffic flow/congestion rate data, then this analysis could be examined to better understand the impact of this program on actual traffic flows. Therefore, we have settled on the SMART question:

_**In the TOLE program, which days of the week earn significantly more or less median daily revenue than the overall median daily revenue?**_

\newpage
## Answering the SMART Question

```{r EDA6, echo=FALSE}
#Calculateand plot the Median Daily Revenue
tickets_day_rev2 <- tickets %>% group_by(Citation.Issue.DayOfWeek) %>% summarise(Median.Daily.Revenue = median(Amount.Paid)*n())

ggplot(data = tickets_day_rev2) + geom_col(mapping = aes(x = tickets_day_rev2$Citation.Issue.DayOfWeek, y= Median.Daily.Revenue), fill="Salmon", color="black") + xlab("Citation Day of Week") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Median Daily Revenue") + ylab("Median Daily Revenue")+ scale_y_continuous(label=comma) + geom_hline(yintercept = median(tickets_day_rev2$Median.Daily.Revenue))


# find median daily revenue by violation type
tickets_viol_median_rev <- tickets %>% group_by(Violation) %>% summarise(Median.Violation.Revenue = median(Amount.Paid)*n()/7)


#Calculateand plot the Median Daily Revenue by Violation type

tickets_day_rev <- tickets %>% group_by(Violation, Citation.Issue.DayOfWeek) %>% summarise(Median.Daily.Revenue = median(Amount.Paid)*n())

ggplot(data = tickets_day_rev) + geom_col(mapping = aes(x = tickets_day_rev$Citation.Issue.DayOfWeek, y= Median.Daily.Revenue), fill="Salmon", color="black") + xlab("Citation Day of Week") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Median Daily Revenue by Violation") + ylab("Median Daily Revenue") + facet_wrap(~Violation) + scale_y_continuous(label=comma)  + geom_hline(data=tickets_viol_median_rev, aes(yintercept = tickets_viol_median_rev$Median.Violation.Revenue))

```

From the "Distribution of Median Daily Revenue," we see that Tuesday appears to generate more revenue than the median daily revenue However, Friday and Saturday appear to have significantly less than median daily revenue. When broken down by violation type in "Distribution of Median Daily Revenue by Violation", Tuesday's high overall value is not able to be visually attributed to any single violation. However, Friday and Saturday can be attributed to below median revenue daily revenue in generation in all violations except `DBL PARK` and `BUS ZONE`, which both seem consistent across all days. Also, `TWAWY ZN#1` has a notably higher than median revenue generation on Monday.



* Variance calculation
* Chi Square 

# NOTES:
  * 

# TO-DO: 
  * CITATIONS!!!!!!!!!!!!!!!!!!!!
  * add summary stats/dimensions to the background of the dataset
  * Variance calcuation
  * Bill introduced in 2007, reevaluated in 2011, made permanent in 2015