---
title: "Project_1_EDA"
author: "Michael Arango, Yeshwant Chillakuru, Jackson Crum"
date: "March 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

----------------------------------------------------------------------

```{r Packages, tidy=T, results='hide', message =F}
#Libraries
library(dplyr)
library(doBy)
library(tidyverse)
library(tidyr)
library(scales)
library(lubridate)
library(ggmap)
```

```{r Read in the file, tidy=T, results = 'hide'}
getwd()
tickets <- data.frame(read.csv("tickets.csv", header=T, sep=','))
```

#Violations

##What TOLE violations are most common?
```{r Most common violations, tidy=T, message=FALSE, echo=FALSE}
table(tickets$Violation)
#Plot violations by count 
ggplot(data=tickets) + geom_bar(mapping = (aes(Violation)), fill = "salmon", color="black") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Number of Tickets by Violation")
```

##What TOLE violations produce the most revenue?
```{r violations most revenue, tidy=T, message=FALSE, echo=FALSE}
#tickets by viol and summed by amount paid
tickets_by_viol <- tickets %>% group_by(Violation) %>% summarise(Total.Revenue = sum(Amount.Paid))
ggplot(tickets_by_viol) + geom_col(mapping = aes(x=Violation, y=Total.Revenue), fill = "salmon", color="black") + scale_y_continuous(labels = comma) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Total Revenue of Tickets by Violation") + ylab("Total Revenue")
  ##Graph for total revenue of violations looks pretty similar to graph of violation counts above

#Remove dataframe
rm(tickets_by_viol)
```

###What TOLE Violations produce the most revenue on weekends and weekdays, on average?
```{r violations most revenue on weekends/weekdays, tidy=T, message=FALSE, echo=FALSE}
#Create variable that stores sum of apount paid by violation and weekday 
tickets.violWeekday <- aggregate(tickets[,c("Amount.Paid")], by=list(tickets$Violation, tickets$Citation.Issue.Weekday), FUN=sum)
  ##Rename the columns of tickets.violWeekday
tickets.violWeekday <- rename(tickets.violWeekday, Violation = Group.1, Weekday = Group.2, Total.Revenue = x)
  ##Create new column for daily average revenue: IF day == weekday, then divide TR by 5; ELSE divide by 2
tickets.violWeekday$Daily.Average.Revenue <- ifelse(tickets.violWeekday$Weekday == 1, tickets.violWeekday$Total.Revenue/5, tickets.violWeekday$Total.Revenue/2)

#Average Weekday and Weekend Revenue by Violation
p <- ggplot(tickets.violWeekday, aes(x=Violation, y=Daily.Average.Revenue, fill=Weekday))
p + geom_col() + coord_flip() + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Total Average Weekday and Weekend Revenue by Violation") + ylab("Daily Average Revenue") + scale_y_continuous(label = comma)

#Remove from Workspace
rm(p, tickets.violWeekday)
```

##For the top 3 violations, what hours are most ticketed? 
```{r top 3 violations most ticketed, tidy=T, message=FALSE, echo=FALSE}
#Figure out which violation types are top 3 most common
ViolCount <- tickets %>% group_by(Violation) %>% summarise(freq=n())
topViol <- ViolCount[order(-ViolCount$freq),]
topViol <- topViol[1:3,]
  ##New dataframe with top 3 violations
tickets.topViol <- subset(tickets, tickets$Violation == topViol$Violation[1] | tickets$Violation == topViol$Violation[2] | tickets$Violation == topViol$Violation[3])
  ##Rename column
tickets.topViol <- rename(tickets.topViol, Top.3.Violations = Violation)
 ##Plot top 3 violations by issue hour
p1 <- ggplot(tickets.topViol, aes(Citation.Issue.Hour, fill=Top.3.Violations))
p1 + geom_density(alpha = .6) + labs(fill = "Top 3 Violations") + ggtitle("Density Distribution of Citation Hour by Top 3 Violations") + theme(plot.title = element_text(hjust = 0.5)) + xlab("Citation Issue Hour")
  ##Generate table of top violation types by citation hour
table(tickets.topViol$Citation.Issue.Hour)
```

##For the top 3 violations, what is the usual amount paid?
```{r top 3 violations amount paid, tidy=T, message=FALSE, echo=FALSE}
p2 <- ggplot(tickets.topViol, aes(Amount.Paid, fill=Top.3.Violations))
p2 + geom_density(alpha = .6) + labs(fill = "Top 3 Violations") + ggtitle("Density Distribution of Amount Paid by Top 3 Violations") + theme(plot.title = element_text(hjust = 0.5)) + xlab("Amount Paid")

#Remove from workspace
rm(p1, p2, topViol, ViolCount, tickets.topViol)
```

##Do TOLE violations vary by time?

###Do TOLE violations vary by hour of the day? 
```{r Violations by hour, tidy=T, message=F}
#Dataframe for number of violations by Hour
Tickets.per.Hour <- tickets %>% group_by(Citation.Issue.Hour) %>% summarise(count = n())
Tickets.per.Hour
  ##Plot violations by hour
ggplot(tickets, aes(x = Citation.Issue.Hour, fill = Violation)) + geom_bar(position = "fill") + scale_x_continuous(breaks = seq(0, 24, 1)) + xlab("Hour of Citation") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Stacked Bar Chart of Violations by Citation Issue Hour ")
# NOTE: The graph shows almost entirely Double Parking and Bus Zone Violations during non-rush-hour and almost entirely Towaway and Prohibited Parking during rush hour

#Remove from workspace
rm(Tickets.per.Hour)
```

```{r Distribution of each Violation by type and hour, tidy=T, message=F, }
#Distribution of Violation by type and hour of day
ggplot(tickets) + geom_histogram(mapping = (aes(x = Citation.Issue.Hour)), fill = "salmon", color = "black", bins = 24) + facet_wrap(~Violation) + xlab("Citation Issue Hour") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Number of Tickets by Violation and Hour of Citation")
```

###Do TOLE violations vary by year? 
```{r Violations by year}
#Violations broken down by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Violation, fill = as.factor(tickets$Citation.Issue.Year)), position = "dodge", color="black") + coord_flip() + xlab("Violation") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Number of Tickets by Violation and Year of Citation") + labs(fill = "Citation Issue Year") 
```

#How does Fine Amount vary?
```{r Fine Amount, tidy=T, message=F}
table(tickets$Fine.Amount)
fine.plot <- ggplot(data = tickets) + geom_histogram(mapping = aes(x = tickets$Fine.Amount), fill = "salmon", color="black")
fine.plot + xlab("Fine Amount") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Histogram of Fine Amount") 
#Because the distribution of fine amount has several high-end outliers, we should be using the median as our measure of central tendency. This is even more clear when comparing the difference between the mean and median (112.7 and 85). 
median(tickets$Fine.Amount)
mean(tickets$Fine.Amount)

  ##By Violation
fine.plot + facet_wrap(~ tickets$Violation) + xlab("Fine Amount") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Histogram of Fine Amount by Violation")
  ## It is clear that tickets have a low fine amount and high fine amount. Is it time of day,      weekday vs weekend, year, etc.? Let's make a factor variable to check this

#Remove from workspace
rm(fine.plot)
```

##There are 2 levels of Fine Amount: Low and High
```{r Fine Level, tidy=T, message=F}
#Initialize new factor variable Fine.Level 
tickets$Fine.Level <- tickets$Fine.Amount
tickets$Fine.Level <- ifelse(tickets$Fine.Amount < 175, "Low", "High")
tickets$Fine.Level <- as.factor(tickets$Fine.Level)
tickets$Fine.Level <- relevel(tickets$Fine.Level, "Low")
  ##Generate table
table(tickets$Fine.Level)
fine.plot2 <- ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Fine.Level), fill = "salmon", color="black")
fine.plot2 + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level") 
  ##Fine level by Violation
fine.plot2 + facet_wrap(~ tickets$Violation) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Violation")

#Remove from workspace
rm(fine.plot2)
```

###Fine Level by Hour of Citation 
```{r Fine level by Hour, tidy=T, message=F}
#Plot Fine level by Hour
fine.plot3 <- ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Hour), fill = "salmon", color="black")
fine.plot3 + facet_wrap(~ tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Hour of Citation")
  ##******High Fine.Level seems to spike at 4 and 5 pm. Is that during Rush Hour? but there is a similar spike in Low Fine.Level too*******

#Remove from workspace
rm(fine.plot3)
```

###Fine Level by weekday and weekend 
```{r Fine level by weekday/weekend, tidy=T, message=F}
#Plot Fine level by weekday vs weekend
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Weekday), fill = "salmon", color="black") + facet_wrap(~ tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Weekday and Weekend")
  ##Does not seem to be a correlation 
```

###Fine Level by day of week
```{r Fine level by day of week, tidy=T, message=F}
#Plot Fine level by day of week
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.DayOfWeek), fill = "salmon", color="black") + facet_wrap(~ tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Day of the Week")
```

###Fine Level by Month
```{r Fine level by month, tidy=T, message=F}
#Plot fine level by month 
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Month), fill = "salmon", color="black") + facet_wrap(~ tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Month")
```

###Fine level by Year
```{r Fine level by year, tidy=T, message=F}
#Plot fine level by year
ggplot(data = tickets) + geom_bar(mapping = aes(x = as.factor(tickets$Citation.Issue.Year)), fill = "salmon", color="black") + facet_wrap(~ tickets$Fine.Level) + xlab("Fine Level") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Fine Level by Year")
```

It looks like they introduced "High" level fines in 2014 as it is the only year in the dataset that a "High" fine level occurs.

#Citation Status
```{r Citation status, tidy=T, message=F}
#Citation Status
table(tickets$Citation.Status)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Status), fill = "salmon", color="black") + xlab("Citation Status") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Citation Status")  
  ##Most are closed. Which ones are still open though?
```

##Citation status by year
```{r Status by year, tidy=T, message=F}
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Status), fill = "salmon", color="black") + xlab("Citation Status") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Bar Chart of Citation Status by Year") + facet_wrap(~ as.factor(tickets$Citation.Issue.Year))
  ##Not make sense to look at fine level or others like that
```

##Citation Status by Amount Due
```{r Status by Amount Due, tidy=T, message=F}
ggplot(data = tickets) + geom_point(mapping = (aes(x = tickets$Citation.Status, y = tickets$Amount.Due)), alpha= 1/50) + xlab("Citation Status") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Dot plot of Citation Status by Amount Due")
```

```{r Status by Amount due 2, tidy=T, message=F}
ggplot(data = tickets) + geom_boxplot(aes(x = tickets$Citation.Status, y = tickets$Amount.Due)) + xlab("Citation Status") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Boxplot of Citation Status by Amount Due") + ylab("Amount Due")
```

###Checking Fine amount vs Amount Due by Citation Status
```{r Amount vs due, tidy =T, message=F}
ggplot(data = tickets) + geom_point(mapping = (aes(x = tickets$Fine.Amount, y = tickets$Amount.Due, color = tickets$Citation.Status))) + xlab("Fine Amount") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Dot plot of Fine Amount vs Amount Due by Citation Status") + ylab("Amount Due")
```

#Amount Paid
```{r Amount paid, tidy=T, message=F}
#Amount paid histogram
ggplot(data = tickets) + geom_histogram(mapping = aes(x = tickets$Amount.Paid), fill = "salmon", color = "black") + xlab("Amount Paid") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Amount Paid") 
#Why are some amounts paid so low? Are these cases where status is open still?
```

##Amount Paid vs Fine Amount
```{r Amount paid vs fine amount, tidy=T, message=F}
#Scatterplot of amount paid and fine amount
ggplot(data = tickets) + geom_point(mapping = aes(x = tickets$Amount.Paid, y=tickets$Fine.Amount, color = tickets$Fine.Level)) + xlab("Amount Paid") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Amount Paid") + labs(color="Fine Level")

#Correlation
cor(tickets$Fine.Amount,tickets$Amount.Paid)
```

There is a surprisingly small and negative correlation between the fine amount and the amount paid. 

##Amount Paid by Citation Status
```{r Amount paid by status, tidy=T, message=F}
tickets_by_paid_status <- tickets %>% group_by(Amount.Paid, Citation.Status) 

ggplot(data = tickets_by_paid_status) + geom_histogram(mapping = aes(x = tickets$Amount.Paid, fill = tickets$Citation.Status), color="black") + xlab("Amount Paid") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Amount Paid by Citation Status") + labs(fill="Citation Status")

#Remove from workspace
rm(tickets_by_paid_status)
```

It makes sense that most of the cases of people paying small amounts or nothing at all are still open. 

# Paid on Time
```{r Paid on Time, tidy=T, message=F}
table(tickets$Paid.On.Time)
ggplot(data = tickets) + geom_bar(mapping = aes(x = as.factor(tickets$Paid.On.Time)), fill="salmon", color="black") + xlab("Paid on Time") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets Paid on Time") 
```

#Time Variables
```{r EDA of Time variables, tidy=T, message=F}
# Month
table(tickets$Citation.Issue.Month)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Month), fill="salmon", color="black") + xlab("Citation Month") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets by Month") 

# Year
table(tickets$Citation.Issue.Year)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Year), fill="salmon", color="black")+ xlab("Citation Year") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets by Year") 

# Day of Week 
table(tickets$Citation.Issue.DayOfWeek)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.DayOfWeek), fill="salmon", color="black") + xlab("Citation Day of Week") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets by Day of Week") 

# Weekday
table(tickets$Citation.Issue.Weekday)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Weekday), fill="salmon", color="black")+ xlab("Citation Issued on Weekend or Weekday") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets by Weekday or Weekend") 

#Hour 
table(tickets$Citation.Issue.Hour)
ggplot(data = tickets) + geom_bar(mapping = aes(x = tickets$Citation.Issue.Hour), fill="salmon", color="black")+ xlab("Citation Hour") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Tickets by Hour") 

# People are getting boned during rush hour on the way home
```

##How have the prices of TOLE violations changed over time?
```{r prices change over time EDA, tidy=T, message=F}

#Calculate the Median Daily Revenue By Violation
tickets_day_rev <- tickets %>% group_by(Violation, Citation.Issue.DayOfWeek) %>% summarise(Median.Daily.Revenue = median(Amount.Paid)*n())
tickets_day_rev
  ##Median Daily Revenue by Violation
ggplot(data = tickets_day_rev) + geom_col(mapping = aes(x = tickets_day_rev$Citation.Issue.DayOfWeek, y= Median.Daily.Revenue), fill="Salmon", color="black") + xlab("Citation Day of Week") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Median Daily Revenue by Violation") + ylab("Median Daily Revenue") + facet_wrap(~Violation) 

##Median Daily Revenue 
tickets_day_rev2 <- tickets %>% group_by(Citation.Issue.DayOfWeek) %>% summarise(Median.Daily.Revenue = median(Amount.Paid)*n())
  ##Plot Median Daily Revenue
ggplot(data = tickets_day_rev2) + geom_col(mapping = aes(x = tickets_day_rev2$Citation.Issue.DayOfWeek, y= Median.Daily.Revenue), fill="Salmon", color="black") + xlab("Citation Day of Week") + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Distribution of Median Daily Revenue") + ylab("Median Daily Revenue") + scale_y_continuous(label=comma)



#median fine amount by violation
fine_by_viol <- tickets %>% 
  group_by(Violation) %>% 
  summarise(median = median(Fine.Amount))
fine_by_viol
  ##Maybe put these results in a table? 

#median fine amount by year
fine_by_year <- tickets %>%
  group_by(Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year)

#Maybe add fine level as control to this*************

#plot median fine amount by year to see trend in prices over time
fine.plot.year <- ggplot(data = fine_by_year, aes(x = Citation.Issue.Year, y = median))
fine.plot.year <- fine.plot.year + geom_line() + geom_point() 
fine.plot.year
# *******Median fine amount per year seems to have increased significantly over time. Index with inflation to see if this is really the case?*******

#fine amount by year, violation
fine_by_year_viol <- tickets %>% 
  group_by(Violation,Citation.Issue.Year) %>% 
  summarise(median = median(Fine.Amount))
head(fine_by_year_viol)

#plot fine by year and violation to see change and trends
fine.plot.yv <- ggplot(data = fine_by_year_viol, aes(x = Citation.Issue.Year, y = median))
fine.plot.yv <- fine.plot.yv + geom_line() + geom_point() 
fine.plot.yv <- fine.plot.yv + facet_wrap(~ Violation)
fine.plot.yv

# ********** Looks like we should look at violations and the years they occur in and maybe get more specific on type of violation*********


rm(fine_by_year, fine_by_viol, fine.plot.year, fine.plot.yv)
```

##Are more citations issued on certain days of the week? Are more tickets issued on weekdays than weekends?

#Jack EDA


```{r Cost Change of Each Violation Type over Hour, Day, Month, and Year}
# These graphs show how Fine Amount changes over each hour, day of the week, month, and year for each Violation Type

ggplot(tickets, aes(x = Citation.Issue.Hour, y = Fine.Amount)) + geom_point() + facet_wrap(~Violation)

ggplot(tickets, aes(x = Citation.Issue.DayOfWeek, y = Fine.Amount)) + geom_point() + facet_wrap(~Violation)

ggplot(tickets, aes(x = Citation.Issue.Month, y = Fine.Amount)) + geom_point() + facet_wrap(~Violation)

ggplot(tickets, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + facet_wrap(~Violation)

# There are appears to be no change over Hour, Day, or Month with a slight increase in prices each year, with a larger spike in 2012. 
```

```{r Cost of six main violations over Year}
#Analysis of price changes over the years by Violation type

#Total of each type of Violation
ViolationCount <- tickets %>% group_by(Violation) %>% summarise(count = n())
ViolationCount

#Create objects for each type of Violation
Violation.DBLPRK <- tickets %>% filter(Violation == "DBL PARK")
Violation.BUSZONE <- tickets %>% filter(Violation == "BUS ZONE")
Violation.PKPHBOTD <- tickets %>% filter(Violation == "PK PHB OTD")
Violation.PRKPROHIB <- tickets %>% filter(Violation == "PRK PROHIB")
Violation.TOWAWAY1 <- tickets %>% filter(Violation == "TWAWY ZN#1")
Violation.TOWAWAYzONE <- tickets %>% filter(Violation == "TWAWY ZONE")

#Graphing the change in fine for each Violation type from 2010-2015

ggplot(Violation.DBLPRK, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Double Parking")

ggplot(Violation.BUSZONE, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Bus Zone")

ggplot(Violation.PKPHBOTD, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Tow-Away Zone - Outside Downtown Core")

ggplot(Violation.PRKPROHIB, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Tow-Away Zone - Downtown Core")

ggplot(Violation.TOWAWAY1, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Tow Away Zone 1")

ggplot(Violation.TOWAWAYzONE, aes(x = Citation.Issue.Year, y = Fine.Amount)) + geom_point() + labs(title = "Tow Away Zone 2")

#General increases in fines over the interval with a noticable jump in prices in 2012
```


## What Hours have the most Violations

```{r Histogram of Citations per Hour, tidy = T}
# Histogram of Citations per Hour
ByHour <- tickets %>% group_by(Citation.Issue.Hour) %>% summarise(Percentage = round(n()/nrow(tickets) * 100, digits = 5))
ByHour

ggplot(tickets, aes(x = Citation.Issue.Hour)) + geom_histogram(binwidth = 1, fill = "red", color = "black") + scale_x_continuous(breaks = seq(0, 24, 1)) 
```

## Are more tickets issued during Rush Hour or Off Hours

```{r}
#Created variable Day of Month for purposes of day-to-day calculations
tickets$Citation.Issue.Day <- day(tickets$Citation.Issue.Date)


RushHour <- tickets %>% filter((Citation.Issue.Hour >= 7 & Citation.Issue.Hour <= 9) | (Citation.Issue.Hour >= 16 & Citation.Issue.Hour <= 18))
NonRushHour <- tickets %>% filter((Citation.Issue.Hour < 7 | Citation.Issue.Hour > 9) & (Citation.Issue.Hour < 16 | Citation.Issue.Hour > 18))
nrow(RushHour)
nrow(NonRushHour)
PercentRush <- nrow(RushHour) / nrow(tickets) * 100
PercentRush
```

```{r}
t <- tickets %>% group_by(Citation.Issue.Date) %>% summarise(median = median(Fine.Amount))
median(t$median)

g <- tickets %>% group_by(Citation.Issue.DayOfWeek) %>% summarise(median = median(Fine.Amount))
g

```


## Comparing Distributions of Rush Hour and Off Hour Totals per Day 

```{r EDA for Rushhour and NonRushhour}
#EDA for Rushhour vs NonRushhour
dim(RushHour)
str(RushHour)
nrow(RushHour)
head(RushHour)

#Calculate total Rushhour violations for each day
Rush <- RushHour %>% group_by(Citation.Issue.Day, Citation.Issue.Month, Citation.Issue.Year) %>% summarise(count = n())
RushTotals <- Rush$count

#EDA of Statistics of Rushhour
mean(RushTotals)
median(RushTotals)
var(RushTotals)
sd(RushTotals)
quantile(RushTotals, c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, .95, 1))
shapiro.test(RushTotals)

#EDA Graphs for Rushhour
qqnorm(RushTotals)
qqline(RushTotals)

boxplot(RushTotals)

plot(RushTotals, dbinom(RushTotals, 30, 0.25), type = "h")

#EDA of NonRushhour
str(NonRushHour)
dim(NonRushHour)
nrow(NonRushHour)

NonRush <- NonRushHour %>% group_by(Citation.Issue.Day, Citation.Issue.Month, Citation.Issue.Year) %>% summarise(count = n())
NonRushTotals <- NonRush$count

mean(NonRushTotals)
median(NonRushTotals)
sd(NonRushTotals)
var(NonRushTotals)
quantile(NonRushTotals, c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, .95, 1))
shapiro.test(NonRushTotals)
qqnorm(NonRushTotals)
qqline(NonRushTotals)

boxplot(NonRushTotals)

ggplot(Rush, aes(x = count)) + geom_histogram(binwidth = 1, fill = "red", color = "black") + scale_x_continuous(breaks = seq(0, 130, 10)) + geom_histogram(data = NonRush, aes(x = count), binwidth = 1, fill = "blue", color = "black", alpha = 0.7) + scale_x_continuous(breaks = seq(0, 130, 10)) 

plot(NonRushTotals, dbinom(NonRushTotals, 30, 0.25), type = "h")

```

## What Violations occur more during Rush Hour and what occur more during Off Hours?

```{r Violation Type per Hour}
# Proportion of Violation Type per Hour

Tickets.per.Hour <- tickets %>% group_by(Citation.Issue.Hour) %>% summarise(count = n())
Tickets.per.Hour

ggplot(tickets, aes(x = Citation.Issue.Hour, fill = Violation)) + geom_bar(position = "fill") + scale_x_continuous(breaks = seq(0, 24, 1)) 

# Graph shows almost entirely Double Parking and Bus Zone Violations during non-rush-hour  and almost entirely Towaway and Prohibited Parking during rush hour

Violation.Rush <- RushHour %>% group_by(Violation) %>% summarise(Total_During_Rush = n())
Violation.Rush

Violation.Off <- NonRushHour %>% group_by(Violation) %>% summarise(Total_During_Off = n())
Violation.Off

a <- Violation.Rush$Total_During_Rush / (Violation.Rush$Total_During_Rush + Violation.Off$Total_During_Off)
#a <- round(a, 3)

```

